An integrated distance sampling approach for modelling age-structured populations

### 2023-06-05 Manuscript_Nilsen_Nater.qmd
Completed draft of model description

The population model follows a post-breeding census and includes two age classes: juveniles (young of the year) and adults (\> 1 year of age, @fig-model). This structure is inspired by earlier models for our focal species, the willow ptarmigan (+ref), and is commonly used for populations of passerine and game birds (+ref). In the context of our willow ptarmigan case study (see below), the census falls into late summer and coincides with the annual distance-sampling survey in August.

@@

$$
\begin{aligned}
  D_{juv, t+1} & = D_{ad,t+1} * R_{t+1} \\
  D_{ad, t+1} & = S_t * (D_{juv,t} + D_{ad, t})  
\end{aligned}
$$ or, alternatively, in matrix notation as

@@

The implementation of the modelling framework we present makes three assumptions about the distance sampling survey: 1) the survey consists of line transects, 2) animals may be detected alone or in groups, and 3) juveniles and adults can be distinguished during surveys. These assumptions are inspired by our willow ptarmigan case study (details below).
Our model includes three likelihoods for different components of the age-structured distance sampling data.

@@

@@

Two different sources of data were used in the open population demographic distance sampling (ODDS) model: i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).

@@

>>>>

The population model follows a post-breeding census and includes two age classes: juveniles (young of the year) and adults (\> 1 year of age, @fig-model). This structure is inspired by earlier models for our focal species, the willow ptarmigan (+ref), and is commonly used for populations of passerine and game birds [@williams2002; @schaub2021]. In the context of our willow ptarmigan case study (see below), the census falls into late summer and coincides with the annual distance-sampling survey in August.

@@

$$
\begin{aligned}
  D_{juv, t+1} & = D_{ad,t+1} * R_{t+1} \\
  D_{ad, t+1} & = S_t * (D_{juv,t} + D_{ad, t})  
\end{aligned}
$$

or, alternatively, in matrix notation as

@@

The implementation of the modelling framework we present makes three assumptions about the distance sampling survey: 1) the survey consists of line transects, 2) animals may be detected alone or in groups, and 3) juveniles and adults can be distinguished during surveys. These assumptions are inspired by our willow ptarmigan case study (details below).
Our model includes three likelihoods for different components of the age-structured distance sampling data. 
First is the likelihood for the perpendicular detection distances from line transect, $y$, which are linked to distance-dependent detection probability $p$ through a half-normal detection function:

$$
\begin{equation}
p = exp(-\frac{y^2}{2\sigma^2})
\end{equation}
$$
where $\sigma$ is the half-normal detection parameter. We assumed $\sigma$ to vary among years (index $t$) but not transect lines or animal group size. Following @moore2011, the resulting $\sigma_t$ can be used to calculate effective strip width ($esw_{t}$) and, consequently, average detection probability per line transect with a truncation distance $W$ according to:

$$
\begin{aligned}
  esw_t & = \sqrt{\frac{\pi * \sigma_{t}^2}{2}}  \\
  \hat{p_t} & = esw_t/W
\end{aligned}
$$ 
The average detection probability $\hat{p_t}$ is an integral part of the second data likelihood which relates the observed number of animals in each age class $a$, $obsN_{a,j,t}$ ($j$ = transect) to the corresponding true number per transect, $N_{a,j,t}$:

$$
\begin{equation}
  obsN_{a,j,t} \sim Poisson(\hat{p_t}*N_{a,j,t})
\end{equation}
$$
$N_{juv,j,t}$ and $N_{ad,j,t}$ are then linked back to the population model by converting them to densities through multiplication with $2L_{j,t}W$ (where $L_{j,t}$ is length of transect $j$ in year $t$).

The third data likelihood focuses on the counts of adults ($obsAd_{j,t}$) and juveniles ($obsJuv_{j,t}$) observed during the distance sampling surveys and links them to time-specific recruitment rate:

$$
\begin{equation}
  obsJuv_{j,t} \sim Poisson(R_t*obsAd_{j,t})
\end{equation}
$$

@@

The final likelihood is for the auxiliary telemetry data. It is set up under the assumption of perfect detection, and hence known fates, of animals bearing transmitters and links the numbers of animals released at the start of season $k$ of year $t$ to the number of survivors at the end of the same season:

$$
\begin{equation}
  survivors_{k,t} \sim Binomial(released_{k,t}, Sk_t)
\end{equation}
$$
Here, $Sk_t$ is the seasonal survival rate, and annual survival rate, $S_t$ is calculated as $S1_t * S2_t$.

#### Hierarchical models with time-variation in parameters

Vital rates (survival probabilities $S$, recruitment rates $R$), detection parameters (half-normal detection parameters $\sigma$), and initial population densities can all be modelled as time-dependent in our framework. For both the tests with simulated data and the case study described below, we implemented log-normally distributed random year effects on all parameters except survival, which was set to be constant. In the case study, we additionally included an effect of rodent occupancy (see details below) on log recruitment rates, resulting in the following model:

$$
\begin{equation}
  log(R_t) = log(\mu_R) + \beta*RodentOcc_t + \epsilon_t
\end{equation}
$$
where $\mu_R$ is the baseline recruitment rate, $\beta$ the slope of the effect of rodent occupancy, and $\epsilon_t$ the normally distributed random effects.

### Model testing with simulated data

TBA

@@

\[Add some basic information about willow ptarmigan in Norway, plus Lierne study area\]

The two datasets used for the case study were i) data from a line transect survey program targeting willow ptarmigan operated as part of the natural resources management authorities, and ii) data from an individual based research project based on radio collared willow ptarmigan. The temporal coverage of the two data sets partially overlap, with the line transect survey spanning 2007-2019 (and ongoing) and the individual based data spanning 2015-2019 (and ongoing).

@@

### Bayesian implementation

We implemented the model in a Bayesian framework using NIMBLE [@valpine2017] in R version 4.2.2 [@R2022]. The likelihood for line transect observation distances was set up using a custom half-normal distribution developed by Michael Scroggie as part of the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance). We used non-informative uniform priors (with biologically reasonable boundaries where possible) for all parameters. For each model fit, we ran x MCMC chains with NIMBLE's standard samples for xxxk iterations. xxxk thereof were discarded as burn-in prior to thining with a factor x, leaving us with xxxx posterior samples per chain (total of xxxxx samples per run).
